An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-09-29 paper.tex
Update 5.2.1

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
% ICWSM 2022 paper Section 3
\subsubsection{Measuring Influence: The IETF as a Social Graph}

\label{subsec:measuring_influence}

Much of the interaction between IETF participants occurs on public mailing
lists.  For each year, we build a social graph based on the \emph{active
community} of participants (nodes) who have interactions (edges) with any
other participant in the previous 5 years.

In Section \ref{sec:org-dyn:participants}, we demonstrated that there
are three categories of participants in IETF - (a) \textit{young
contributors} -- who leave within one year of their first year mailing list
contribution; (b) \textit{mid-age contributors} -- participants who stay
active for up-to 5 years; and (c) \textit{senior contributors} -- who
remain active for more than 5 years. Following this, a 5 year period window
is chosen to observe interactions. There are no direct
participant-to-participant email exchanges in our dataset: the emails and
responses we capture are sent to the public mailing lists, and we observe
an interaction between two participants when one replies to an email sent
by another on any mailing list. This yields a social graph based on
1,049,793 emails (out of 2.1M) from 22,138 unique participants across 840
mailing lists.

%..................................................................................................
% ICWSM 2022 paper Section 3.1
\pb{Participation}
To examine the variation in participation between IETF participants, in
Figure~\ref{fig:degree_cdf} we plot the cumulative degree distribution of
the social graph -- i.e., the cumulative number of people a participant
interacts with.  We observe a core group that always interacts with
substantially more people than the rest of the community, with around 80\%
of the participants having degree  $<$15, but 5\%-10\% having degree $>$40.
However,  this difference decreases over time: in 2005-2009 around 6\% of
participants had degree $>$40, but by the period 2015-2019 this increases
to $\sim10\%$.  This shows that while there is always a core of more active
participants, the degree of participation has spread  out over time.
\todo{Repetitive with Section \ref{sec:org-dyn:participants}?}

@@

To understand the structure of the community, and its reliance on specific
groups, we analyse the connected components of this graph. Each connected
component reflects a maximal set of nodes such that each pair of nodes is
connected by a path. We compute the size of the Largest Connected Component
(LCC) and the Number of Connected Components (NCC) for each year in
Figure~\ref{fig:lcc_size_yearly}.  The NCC peaks in 2003 before declining,
and broadly aligns with the variation in the number of meeting
attendees,\footnote{\url{https://datatracker.ietf.org/stats/meeting/overview/}}
suggesting a lag between participating for the first time and integrating
into the wider community.  In contrast, the size of the LCC increases until
2006 before stabilising.  Overall, we observe that the IETF community has
become less fragmented.

\begin{figure}
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lcc_yearly_no_components.pdf}
  \caption{
    Size of Largest Connected Component (LCC) and Number of Connected
    Components (NCC)
  }
  \label{fig:lcc_size_yearly}
\end{figure}

%..................................................................................................
% ICWSM 2022 paper Section 3.1
\pb{Influence}
Relying on a small group to interconnect the community could undermine the
resilience of the IETF.  To study the influence of participants and their
role in interconnecting the wider community, we compute the betweenness
centrality of each participant in a given time
period~\cite{kourtellis2013identifying, weitzel2012measuring,
sole2014centrality}. We also considered other graph based influence metrics
such as eigenvector centrality: eigenvector centrality reflects the
importance of a node as per its neighbours, while betweenness centrality is
based on shortest paths which is independent of the influence of
neighbours.

However, we found a very strong correlation between the two measures, 
in line with similar experiments \cite{valente2008correlated,he2016correlation}:
the Spearman's rank correlation of participants ranked by betweenness
centrality and eigenvector centrality ranged between 0.51-0.72, with a
strong statistical significance ($p < 0.01$) for the period 2000-2019. We
therefore use just betweenness centrality in our analysis; this is often
studied and acknowledged as a measure of influence in social and complex
networks, particularly built on online communication
\cite{hagen2018crisis,chen2012identifying, ghalmane2019centrality}.

Figure~\ref{fig:purge_nodes_largest_connected_component_number_components}
then shows the effect of removing the most influential participants (in 1\%
increments from most to least, moving left-to-right on the x-axis) on the
size of the LCC.  Worryingly, we find that removal of just 20-25\% of the
most influential participants causes the LCC to shrink by 90\%.  However,
we also find that this impact has decreased over time: for instance, in
2000-2004, removing the top $\sim$5\% influential participants, reduces the
size of LCC by more than half, whereas in 2015-2019, it takes the removal
of the top $\sim$15\% of participants to have the same effect.  This shows
that the community has become \emph{more} cohesive and resilient over time,
and the IETF can now sustain a larger amount of churn while maintaining a
well-connected social graph.

>>>>

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsubsection{Measuring Influence: The IETF as a Social Graph}
\label{subsec:measuring_influence}

% ICWSM 2022 paper Section 3

To measure influence of participants in the IETF we extend the analysis in
Section \ref{sec:org-dyn:participants} and, for each year, build a social
graph based on the \emph{active community} of participants (nodes) who have
email interactions (edges) with any other participant in the previous five
years.

We previously demonstrated that there are three categories of participants
in IETF: \textit{young contributors} who leave within one year of their
first year mailing list contribution; \textit{mid-age contributors} who
stay active for up-to five years; and \textit{senior contributors} who
remain active for more than five years (\S\ref{sec:org-dyn:participants}).
Following this, a 5-year period window is chosen to observe interactions.
There are no direct participant-to-participant email exchanges in our
dataset: the emails and responses we capture are sent to the public mailing
lists, and we observe an interaction between two participants when one
replies to an email sent by another on any mailing list. This yields a
social graph based on 1,049,793 emails (out of 2.1M; not all messages
receive a reply) from 22,138 unique participants across 840 mailing lists.

%..................................................................................................
% ICWSM 2022 paper Section 3.1
\pb{Participation}
To examine the variation in participation between IETF participants, in
Figure~\ref{fig:degree_cdf} we plot the cumulative degree distribution
of the social graph (i.e., the cumulative number of people with which a
participant interacts). We observe a core group that always interacts with
substantially more people than the rest of the community, with around 80\%
of the participants interacting with less than 15 others (i.e., having
degree~$<15$), but 5\%-10\% interacting with more than 40 others.
This difference decreases over time: in 2005-2009 only around 6\% of
participants had degree $>$40, but by the period 2015-2019 this increases
to around 10\%.
While there is always a core of more active participants, the amount of
interaction has increased over time.
This confirms the results in Figure \ref{fig:degree_dist_rfc_authors},
for RFC authors, and shows that they also apply to interactions in the
community at large.

@@

To understand the structure of the community, and its reliance on specific
groups, we analyse the connected components of this graph. Each connected
component reflects a maximal set of nodes such that each pair of nodes is
connected by a path. We compute the size of the Largest Connected Component
(LCC) and the Number of Connected Components (NCC) for each year in
Figure~\ref{fig:lcc_size_yearly}.  The NCC peaks in 2003 before declining,
and broadly aligns with the variation in the number of meeting
attendees,\footnote{\url{https://datatracker.ietf.org/stats/meeting/overview/}}
suggesting a lag between participating for the first time and integrating
into the wider community.  In contrast, the size of the LCC increases until
2006 before stabilising.  Overall, we observe that the IETF community has
become less fragmented over time.

\begin{figure}
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2022/lcc_yearly_no_components.pdf}
  \caption{
    Size of Largest Connected Component (LCC) and Number of Connected
    Components (NCC) of the email graph.
  }
  \label{fig:lcc_size_yearly}
\end{figure}

%..................................................................................................
% ICWSM 2022 paper Section 3.1
\pb{Influence}
Relying on a small group to interconnect the community could undermine the
resilience of the IETF.  To study the influence of participants and their
role in interconnecting the wider community, we compute the betweenness
centrality of each participant in a given time
period~\cite{kourtellis2013identifying, weitzel2012measuring,
sole2014centrality}.%
\footnote{
  We considered other graph based influence metrics such as eigenvector
  centrality: eigenvector centrality reflects the importance of a node as
  per its neighbours, while betweenness centrality is based on shortest
  paths which is independent of the influence of neighbours.  However, we
  found a very strong correlation between the two measures, in line with
  similar experiments \cite{valente2008correlated,he2016correlation}: the
  Spearman's rank correlation of participants ranked by betweenness
  centrality and eigenvector centrality ranged between 0.51-0.72, with a
  strong statistical significance ($p < 0.01$) for the period 2000-2019. We
  therefore use just betweenness centrality in our analysis; this is often
  studied and acknowledged as a measure of influence in social and complex
  networks, particularly built on online communication
  \cite{hagen2018crisis,chen2012identifying, ghalmane2019centrality}.
}
Figure~\ref{fig:purge_nodes_largest_connected_component_number_components}
then shows the effect of removing the most influential participants (in 1\%
increments from most to least, moving left-to-right on the x-axis) on the
size of the LCC.  Worryingly, we find that removal of just 20-25\% of the
most influential participants causes the LCC to shrink by 90\%.  However,
we also find that this impact has decreased over time: for instance, in
2000-2004, removing the top $\sim$5\% influential participants, reduces the
size of LCC by more than half, whereas in 2015-2019, it takes the removal
of the top $\sim$15\% of participants to have the same effect.  This shows
that the community has become \emph{more} cohesive and resilient over time,
and the IETF can now sustain a larger amount of churn while maintaining a
well-connected social graph.
